---
title: "Predicting the Distribution of Gastroenteritis-Causing Agents"
author: "Antonette Tadle"
date: "2023-12-22"
output: pdf_document
---

I. Introduction

# Ecologically, rivers provide significant habitats for a range of plants and wildlife. However, they are also prone to pollution due to a variety of human-induced hazards such as effluents from industries and sewage discharge, runoffs, lack of treatment and sewage systems, sedimentation, and rapid urbanization (Lin et al., 2022). The Butuanon River is a 23-kilometer river nestled between Mandaue and Cebu City that is defined as Cebu’s “dead river” because of extreme pollution which hinders floral and faunal growth (Parilla et al., 2019). High levels of contaminants, such as industrial effluents and untreated sewage, disrupt the river's ecosystem. This pollution can lead to a decline in aquatic plant populations, affecting the oxygen levels and nutrient cycling essential for a healthy ecosystem (Garg et al., 2022). Additionally, it poses a threat to various fauna species by contaminating their habitats, disrupting food chains, and causing long-term harm to biodiversity (Kaur and Brraich, 2022). Being one of the seven major rivers of Cebu and as a crucial water source, Butuanon river plays a vital role in the lives of the local population, making it imperative to understand the epidemiological drivers of water-borne diseases in this context. Discerning the environmental factors supporting the presence of a disease agent elucidates the disease's ecology and geography (Peterson, 2007). 

# Gastroenteritis is among the highest causes of morbidity in Cebu province and is the one of the main causes of infant mortality. It is associated with a variety of identified pathogens in the Butuanon river including Salmonella, Enterococcus, Enterobacter, E. coli, Klebsiella, Citrobacter, Serratia, Pseudomonas, Proteus, Providencia, and Acinetobacter which causes viral diarrhea. Short-term health responses to water quality investigations have revealed effects on infant mortality and gastroenteritis incidence related to biological contamination. These responses are significantly connected because fecal coliform contamination resulting in repeated bouts of diarrhea has been proven to impair the child's ability to absorb nutrition, leading to mortality or growth stunting (Gadgil, 1998). 

# Total coliforms, a group of bacteria found in the environment, are widely used as indicators of water quality, especially in assessing the safety of drinking water and recreational waters. They include a range of bacteria, some of which originate from the intestines of warm-blooded animals, including humans. Their presence in water indicates fecal contamination and the potential presence of pathogens that cause gastroenteritis, such as E. coli, Salmonella, and Shigella. Several studies have established a correlation between elevated total coliform counts in water sources and increased cases of gastroenteritis within affected communities. However, variability in coliform levels due to seasonal changes, geographical factors, and human activities requires sophisticated modeling techniques for accurate predictions.

# Currently, disease biogeography is poorly addressed in biodiversity research (Peterson et al., 2011). While substantial research has been conducted on the distribution of species and their interactions in ecosystems, the study of diseases within this context has received insufficient attention. As a result, there is a substantial gap in our understanding of how illnesses spread across different geographical regions and interact with biodiversity patterns. Species distribution modeling is used in epidemiology for mapping spatial disease patterns, the prediction of disease introduction risks through pathogen-host interactions, and the prediction of exposure changes owing to future environmental changes (Martínez-Minaya et al., 2018). While historical records of infections and vectors were limited, the introduction of monitoring systems and media sources has resulted in new online data sources on their occurrences. Furthermore, distribution modeling tools are becoming more ecologically realistic by taking dispersal, biotic interactions, and evolutionary restrictions into account. These factors, together with abiotic circumstances and recording biases, jointly influence disease, vector, and wildlife species distributions. 

# Ecological niche modeling (ENM) is an advanced computational approach that is used to explain the occurrence of infectious agents with the influence of environmental factors (Escobar and Craft, 2016).  Specifically, correlative ENM is a convenient approach in predicting site-specific biogeography since they only require environmental data associated with species occurrence locations to predict the spatial distribution of species (Graham et al., 2004; Guisan and Thuiller, 2005, as cited in Sebes, 2023). It is preferred over mechanistic ENM modeling because it can capture environmental and socioeconomic factors for predicting and understanding patterns in transmission (Hay et al., 2013, as cited in Purse and Golding, 2015). Consequently, species distribution modeling (SDM) is often associated with ENM to correlate species occurrence data with environmental data, such as MaxEnt (Phillips et al., 2006, as cited in Sebes, 2023). These modeling approaches are supplemented with GAMs and ANNs for generating nonlinear statistical models, species distribution modeling, environmental monitoring, and image classification; GCMs to model the spread of species and disease changes over time under different future climatic conditions; and Ensemble to combine multiple algorithms or models to improve predictive accuracy and robustness. 

# This study addresses diverse challenges related to these approaches including better data organization, such as quantification of pixel count changes and modeling at increased resolutions. Applying correlative ENM and SDM to the context of water-borne diseases provides a novel perspective on disease ecology. This study can identify pollution sources and vulnerable areas by generating species distribution maps of gastroenteritis-causing pathogens and physicochemical factors. Furthermore, the study’s findings can inform targeted interventions and regulations for pollution sources, recommend effective water treatment methods, assess health risks associated with pathogens, promote community awareness for water safety, and contribute to evidence-based policymaking and climate change adaptation for better public health outcomes.

II. Data
# Existing data from Project REHAB containing identified species and physicochemical data in Butuanon River water samples was used as input data points in this project. This data was obtained from sampling periods in July and September 2023 and was wrangled and transformed into a uniform format. The data includes latitude and longitude values for both species occurrence and environmental data, as well as values for each physicochemical parameter including temperature, oxidation-reaction potential, pH, DOP, DO, EC, TDS, salinity, BGA-PC, altitude, and barometer. 
```{r}
library(readxl)
excel_file <- read_excel("~/Desktop/bio 118-class/exercises/bio 118 project/bio-118-project-attadle/Data/combined_data.xlsx")
library(dplyr)
glimpse(excel_file)
head(excel_file)
str(excel_file)
```

III. Data Analysis
# The central question in this study is to assess the distribution of gastroenteritis-causing agents by correlating the spatial distribution of species with physicochemical data. 

IV. Results and Discussion
# The data was filtered to include only the Gastroenteritis-causing pathogens in a species distribution map. The map features longitude values on the x-axis and latitude values on the y-axis.
```{r}
# load the required libraries
library(tidyverse)
library(sp)
library(leaflet)
library(readxl)
library(ggplot2)
library(dplyr)
library(readxl)
library(mapview)
library(spdep)
library(patchwork)
library(sf)

# Species Occurrence Distribution Map
excel_file <- "./Data/combined_data.xlsx" 
data <- read_excel(excel_file, sheet = "1. Species Occurrence")
selected_species <- c("E. coli", "Salmonella", "Shigella", "V. cholerae", "V. fluvialis", "V. parahaemolyticus", "V. vulnificus", "Aeromonas")

# Filter data
filtered_data1 <- data %>%
  filter(Species %in% selected_species, 
         !Species %in% c("Unknown", "Unknown (Vibrio)")) %>%
  select(-"Concentration")

# Convert to a regular data frame
filtered_data_df <- fortify(filtered_data1, region = "Species")

# Create an sp object
coordinates(filtered_data1) <- c("Lon", "Lat")

# River Shape Files
rivers <- sf::read_sf("./Data/Shape_Files/QGIS files/Sampling Sites River Path/Sampling Sites River Path.shp") 
ph_shp <- sf::read_sf("./Data/Shape_Files/QGIS files/Philippine Map/PHL_adm/PHL_adm3.shp")

# Plot the map
ggplot(data = filtered_data_df) +
  geom_sf(data = ph_shp) +
  geom_sf(data = rivers) +
  coord_sf(xlim = c(123.85, 124), ylim = c(10.33, 10.45), expand = FALSE) +
  geom_point(aes(x = Lon, y = Lat, color = Species), size = 2) +
  theme_bw()
```
# The graph shows the distribution of specific gastroenteritis-causing agents along the Butuanon river including Aeromonas, E. coli, Salmonella, Shigella, V. cholerae, V. fluvialis, V. parahaemolyticus, and V. vulnificus. It is evident that some points in the graph did not fall exactly within the Butuanon River course due to errors during the first sampling period. It can be inferred that E. coli had the largest distribution across all sites from upstream to downstream, followed by Salmonella, V. cholerae, and Shigella. Meanwhile, V. fluvialis, V. parahaemolyticus, V. vulnificus, and aeromonas had the least distribution across all sites.

# For every physicochemical parameter, a distribution map was created, depicting longitude values on the x-axis and latitude values on the y-axis. Each parameter value was assigned a color gradient for visual representation.
```{r}
library(leaflet)
library(mapview)
library(htmlwidgets)
library(readxl)
library(ggplot2)
library(dplyr)
library(patchwork)
getwd()

# Read data
excel_file <- "./Data/combined_data.xlsx" 
data2 <- read_excel(excel_file, sheet = "2. Environmental Factors")
data2$Parameter <- as.character(data2$Parameter)

# Temperature
temp <- dplyr::filter(data2, Parameter == "Temperature")
plot_temp <- ggplot(data = data2)+
  geom_sf(data = ph_shp)+
  geom_sf(data = rivers)+
  geom_tile(data = temp, aes(x = Lon, y = Lat, fill = Value), height = .005, width = .005)+
  coord_sf(xlim = c(123.85, 124), ylim = c(10.33, 10.45), expand = FALSE)+
  #geom_point(aes(x=Lon, y=Lat, color= Species), size= 2)+
  theme_bw()+
  ggtitle("Temperature (in Celcius)")

ggsave(filename = paste0("./Outputs/", "Temperature.png"), device = "png")

# Oxidation-Reaction Potential
OR <- dplyr::filter(data2, Parameter == "Oxidation-Reaction Potential")
plot_OR <- ggplot(data = data2)+
  geom_sf(data = ph_shp)+
  geom_sf(data = rivers)+
  geom_tile(data = OR, aes(x = Lon, y = Lat, fill = Value), height = .005, width = .005)+
  coord_sf(xlim = c(123.85, 124), ylim = c(10.33, 10.45), expand = FALSE)+
  #geom_point(aes(x=Lon, y=Lat, color= Species), size= 2)+
  theme_bw()+
  ggtitle("Oxidation-Reaction Potential (mV)")

# pH
pH <- dplyr::filter(data2, Parameter == "pH")
plot_pH <- ggplot(data = data2)+
  geom_sf(data = ph_shp)+
  geom_sf(data = rivers)+
  geom_tile(data = pH, aes(x = Lon, y = Lat, fill = Value), height = .005, width = .005)+
  coord_sf(xlim = c(123.85, 124), ylim = c(10.33, 10.45), expand = FALSE)+
  #geom_point(aes(x=Lon, y=Lat, color= Species), size= 2)+
  theme_bw()+
  ggtitle("pH")

# DO(mg/L)
DO <- dplyr::filter(data2, Parameter == "DO")
plot_DO <- ggplot(data = data2)+
  geom_sf(data = ph_shp)+
  geom_sf(data = rivers)+
  geom_tile(data = DO, aes(x = Lon, y = Lat, fill = Value), height = .005, width = .005)+
  coord_sf(xlim = c(123.85, 124), ylim = c(10.33, 10.45), expand = FALSE)+
  #geom_point(aes(x=Lon, y=Lat, color= Species), size= 2)+
  theme_bw()+
  ggtitle("DO (mg/L)")

# EC
EC <- dplyr::filter(data2, Parameter == "EC")
plot_EC <- ggplot(data = data2)+
  geom_sf(data = ph_shp)+
  geom_sf(data = rivers)+
  geom_tile(data = EC, aes(x = Lon, y = Lat, fill = Value), height = .005, width = .005)+
  coord_sf(xlim = c(123.85, 124), ylim = c(10.33, 10.45), expand = FALSE)+
  #geom_point(aes(x=Lon, y=Lat, color= Species), size= 2)+
  theme_bw()+
  ggtitle("Electrical Conductivity (uS/cm)")

# TDS
TDS <- dplyr::filter(data2, Parameter == "TDS")
plot_TDS <-ggplot(data = data2)+
  geom_sf(data = ph_shp)+
  geom_sf(data = rivers)+
  geom_tile(data = TDS, aes(x = Lon, y = Lat, fill = Value), height = .005, width = .005)+
  coord_sf(xlim = c(123.85, 124), ylim = c(10.33, 10.45), expand = FALSE)+
  #geom_point(aes(x=Lon, y=Lat, color= Species), size= 2)+
  theme_bw()+
  ggtitle("Total Dissolved Solids (mg/L)")

# Sal
sal <- dplyr::filter(data2, Parameter == "SAL")
plot_sal <- ggplot(data = data2)+
  geom_sf(data = ph_shp)+
  geom_sf(data = rivers)+
  geom_tile(data = sal, aes(x = Lon, y = Lat, fill = Value), height = .005, width = .005)+
  coord_sf(xlim = c(123.85, 124), ylim = c(10.33, 10.45), expand = FALSE)+
  #geom_point(aes(x=Lon, y=Lat, color= Species), size= 2)+
  theme_bw()+
  ggtitle("Salinity (PSU)")

# BGA-PC
BGAPC <- dplyr::filter(data2, Parameter == "BGA-PC")
plot_BGAPC <- ggplot(data = data2)+
  geom_sf(data = ph_shp)+
  geom_sf(data = rivers)+
  geom_tile(data = BGAPC, aes(x = Lon, y = Lat, fill = Value), height = .005, width = .005)+
  coord_sf(xlim = c(123.85, 124), ylim = c(10.33, 10.45), expand = FALSE)+
  #geom_point(aes(x=Lon, y=Lat, color= Species), size= 2)+
  theme_bw()+
  ggtitle("Phycocyanin Blue-Green Algae Sensor (nm)")

# Altitude
alt <- dplyr::filter(data2, Parameter == "Altitude")
plot_alt <- ggplot(data = data2)+
  geom_sf(data = ph_shp)+
  geom_sf(data = rivers)+
  geom_tile(data = temp, aes(x = Lon, y = Lat, fill = Value), height = .005, width = .005)+
  coord_sf(xlim = c(123.85, 124), ylim = c(10.33, 10.45), expand = FALSE)+
  #geom_point(aes(x=Lon, y=Lat, color= Species), size= 2)+
  theme_bw()+
  ggtitle("Altitude (PSI)")

# Barometer
baro <- dplyr::filter(data2, Parameter == "Barometer")
plot_baro <-ggplot(data = data2)+
  geom_sf(data = ph_shp)+
  geom_sf(data = rivers)+
  geom_tile(data = baro, aes(x = Lon, y = Lat, fill = Value), height = .005, width = .005)+
  coord_sf(xlim = c(123.85, 124), ylim = c(10.33, 10.45), expand = FALSE)+
  #geom_point(aes(x=Lon, y=Lat, color= Species), size= 2)+
  theme_bw()+
  ggtitle("Barometer (PSI)")

# Depth
depth <- dplyr::filter(data2, Parameter == "Depth")
plot_depth <-ggplot(data = data2)+
  geom_sf(data = ph_shp)+
  geom_sf(data = rivers)+
  geom_tile(data = depth, aes(x = Lon, y = Lat, fill = Value), height = .005, width = .005)+
  coord_sf(xlim = c(123.85, 124), ylim = c(10.33, 10.45), expand = FALSE)+
  #geom_point(aes(x=Lon, y=Lat, color= Species), size= 2)+
  theme_bw()+
  ggtitle("Depth (m)")

combined_plot1 <- plot_temp + plot_OR + plot_pH + plot_DO + plot_EC + plot_TDS

combined_plot2 <- plot_sal + plot_BGAPC + plot_alt + plot_baro + plot_depth
  
combined_plot1
combined_plot2
```
# The observed species occurrence data reveals interesting correlations with the visualized physicochemical parameters across the sampled sites. E. coli exhibits the broadest distribution, spanning from upstream to downstream, suggesting its potential in surviving through various environments and a broader environmental niche. Salmonella also demonstrates a widespread presence, albeit less extensive than E. coli. On the other hand, V. cholerae and Shigella show narrower distributions, hinting at specific ecological niches or preferences within the water system. The lower distribution of V. fluvialis, V. parahaemolyticus, V. vulnificus, and aeromonas across all sites may indicate a more localized or selective habitat for these species.

# Examining the physicochemical parameters, variations in barometric pressure and altitude reveal distinct patterns. Barometer values increase downstream, suggesting potential atmospheric influences, while higher altitude in mid to downstream areas indicates topographical changes. Depth peaks at Site 14 in the bay, reflecting the specific hydrological conditions of this location. BGA-PC shows variability across sites, with the highest concentrations in the lower midstream areas at sites 8 and 9, indicating potential nutrient enrichment. Salinity also increases downstream, possibly influencing the distribution of certain species.

# Temperature exhibits a gradient with higher values in mid to downstream regions, aligning with the broader distribution of E. coli and Salmonella. Oxidation-reduction potential (OR) is highest upstream and midstream, potentially influencing the occurrence of specific species adapted to these conditions. pH and dissolved oxygen levels are greatest upstream, suggesting a preference for more oxygenated and neutral environments by certain species. Electrical conductivity peaks downstream, possibly influencing the distribution of species adapted to higher conductivity levels. Total dissolved solids (TDS) also show an increasing trend downstream, which may impact the habitat preferences of certain species..




